---
title: "Time Allocation Patterns and Well-Being"
author: "Sashwat Desai"
date: "2025-02-25"
output: html_document
---

```{r Clear Environment}
rm(list = ls())
```

## Loading the Necessary Libraries

```{r Libraries}
library(tidyverse)      # Data manipulation and visualization
library(ggplot2)        # Advanced plotting
library(corrplot)       # Correlation visualization
library(car)            # For VIF calculations
library(vcd)            # For categorical data visualization
library(knitr)          # For nice tables
library(gridExtra)      # For arranging multiple plots
```

## [1] DATA LOADING AND PREPARATION

```{r Data Loading and Preparation}
# Load the data
covid_data <- read.csv("COVID-19 Survey Student Responses.csv")

# First, let's examine the actual column names to ensure we're using the right ones
print(names(covid_data))

# Inspect the first few rows to understand the data structure
head(covid_data)

# Clean approach - create a more general function to find columns by partial name
find_column <- function(df, pattern) {
  cols <- grep(pattern, names(df), ignore.case = TRUE, value = TRUE)
  if(length(cols) == 0) {
    warning(paste("No column matching pattern:", pattern))
    return(NULL)
  }
  return(cols[1])  # Return the first matching column
}

# Find health and weight change columns
health_col <- find_column(covid_data, "health|issue")
weight_col <- find_column(covid_data, "weight|change")
meals_col <- find_column(covid_data, "meal")

# Display the found column names
cat("Health column found:", health_col, "\n")
cat("Weight column found:", weight_col, "\n")
cat("Meals column found:", meals_col, "\n")

# Create binary health issue variable more robustly
if(!is.null(health_col)) {
  # First check the unique values to better understand the data
  print(table(covid_data[[health_col]]))
  
  # Create binary health issue indicator (Yes/No) - handles various "no" formats
  covid_data$has_health_issue <- ifelse(
    grepl("^[Nn][Oo]$|^[Nn][Oo][Nn][Ee]$", covid_data[[health_col]]) | 
      is.na(covid_data[[health_col]]) | 
      covid_data[[health_col]] == "", 
    "No", "Yes")
  
  # Convert to factor
  covid_data$has_health_issue <- factor(covid_data$has_health_issue)
} else {
  # Create a default if column not found
  covid_data$has_health_issue <- factor("Unknown")
  warning("Health issue column not found, using placeholder")
}

# Handle weight change column
if(!is.null(weight_col)) {
  # Convert to factor
  covid_data$weight_change <- factor(covid_data[[weight_col]])
} else {
  covid_data$weight_change <- factor("Unknown")
  warning("Weight change column not found, using placeholder")
}

# Find time variables - look for columns with "time" and specific activities
sleep_col <- find_column(covid_data, "sleep")
fitness_col <- find_column(covid_data, "fitness")
social_media_col <- find_column(covid_data, "social media")
tv_col <- find_column(covid_data, "tv")

# Display found time columns
cat("Sleep column found:", sleep_col, "\n")
cat("Fitness column found:", fitness_col, "\n")
cat("Social media column found:", social_media_col, "\n")
cat("TV column found:", tv_col, "\n")

# Create a vector of found time columns
time_vars <- c(sleep_col, fitness_col, social_media_col, tv_col)
time_vars <- time_vars[!is.null(time_vars)]  # Remove any NULL values

# Convert time variables to numeric
for(var in time_vars) {
  cat("Processing column:", var, "\n")
  # Handle different formats and convert to numeric
  covid_data[[var]] <- as.numeric(as.character(covid_data[[var]]))
  # Print summary to check
  print(summary(covid_data[[var]]))
}

# Create dataset for analysis, dropping rows with missing values in key variables
analysis_cols <- c("has_health_issue", "weight_change", time_vars)
if(!is.null(meals_col)) {
  analysis_cols <- c(analysis_cols, meals_col)
}

# Print the columns we'll use for analysis
cat("Columns for analysis:", paste(analysis_cols, collapse=", "), "\n")

analysis_data <- covid_data %>%
  select(all_of(analysis_cols)) %>%
  drop_na()

# Print dimensions to verify we have data
cat("Analysis dataset dimensions:", dim(analysis_data)[1], "rows,", 
    dim(analysis_data)[2], "columns\n")
```

## [2] EXPLORATORY DATA ANALYSIS

```{r Exploratory Data Analysis}
# Summary statistics by health issue status
time_by_health <- analysis_data %>%
  group_by(has_health_issue) %>%
  summarise(across(all_of(time_vars), 
                  list(mean = ~mean(., na.rm = TRUE),
                       sd = ~sd(., na.rm = TRUE),
                       median = ~median(., na.rm = TRUE))))

print(time_by_health)

# Summary statistics by weight change
time_by_weight <- analysis_data %>%
  group_by(weight_change) %>%
  summarise(across(all_of(time_vars), 
                  list(mean = ~mean(., na.rm = TRUE),
                       sd = ~sd(., na.rm = TRUE),
                       median = ~median(., na.rm = TRUE))))

print(time_by_weight)

# Boxplots for time variables by health issue
health_plots <- list()
for(var in time_vars) {
  p <- ggplot(analysis_data, aes_string(x = "has_health_issue", y = var, fill = "has_health_issue")) +
    geom_boxplot() +
    theme_minimal() +
    labs(title = paste("Distribution of", gsub("\\.", " ", var), "by Health Issue"),
         x = "Reported Health Issue", y = "Hours per Day") +
    theme(legend.position = "none")
  health_plots[[var]] <- p
}

# Arrange multiple boxplots in a grid
do.call(grid.arrange, c(health_plots, ncol = 2))

# Correlation matrix of time variables
cor_matrix <- cor(analysis_data[, time_vars], use = "complete.obs")
corrplot(cor_matrix, method = "circle", type = "upper", 
         title = "Correlation Between Time Allocation Variables")
```

## [3] STATISTICAL TESTS

(a) Chi-Square Test for Association Between Weight Change and Health Issues

```{r Chi-Square Test for Association Between Weight Change and Health Issues}
calculate_cramers_v <- function(x, y) {
  tbl <- table(x, y)
  chi2 <- chisq.test(tbl)$statistic
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  sqrt(chi2 / (n * min_dim))
}

chisq_result <- chisq.test(table(analysis_data$weight_change, analysis_data$has_health_issue))
print(chisq_result)

# Calculate Cramer's V for effect size using our custom function
cramer_v <- calculate_cramers_v(analysis_data$weight_change, analysis_data$has_health_issue)
cat("Cramer's V:", cramer_v, "\n")

# Visualize the association
if(require(vcd)) {
  mosaic(~ weight_change + has_health_issue, data = analysis_data,
         main = "Association Between Weight Change and Health Issues")
} else {
  # Alternative visualization if vcd package is not available
  ggplot(analysis_data, aes(x = weight_change, fill = has_health_issue)) +
    geom_bar(position = "fill") +
    theme_minimal() +
    labs(title = "Association Between Weight Change and Health Issues",
         x = "Weight Change", y = "Proportion",
         fill = "Health Issue Reported")
}
```

(b) ANOVA: Compare Time Spent on Activities Between Health Issue Groups

```{r ANOVA: Compare Time Spent on Activities Between Health Issue Groups}
anova_results <- list()
for(var in time_vars) {
  formula <- as.formula(paste(var, "~ has_health_issue"))
  anova_result <- aov(formula, data = analysis_data)
  anova_results[[var]] <- anova_result
  cat("\nANOVA for", gsub("\\.", " ", var), "by Health Issue:\n")
  print(summary(anova_result))
}
```

(c) Logistic Regression: Predict Health Issues From Time Allocation

```{r Logistic Regression: Predict Health Issues From Time Allocation}
# Update variable names to match the dataset
logistic_model <- glm(has_health_issue ~ Time.spent.on.sleep + Time.spent.on.fitness + 
                              Time.spent.on.TV + Number.of.meals.per.day,
                    family = binomial(link = "logit"), 
                    data = analysis_data)

summary(logistic_model)

# Check for multicollinearity
vif(logistic_model)

# Calculate odds ratios
odds_ratios <- exp(coef(logistic_model))
# Use tryCatch to handle potential errors in confint
or_ci <- tryCatch({
  exp(confint(logistic_model))
}, error = function(e) {
  # Use profile likelihood method if default method fails
  exp(confint.default(logistic_model))
})

odds_ratio_table <- data.frame(
  OR = odds_ratios,
  Lower_CI = or_ci[,1],
  Upper_CI = or_ci[,2]
)
print(odds_ratio_table)
```

(d) Multinomial Logistic Regression for Weight Change

```{r Multinomial Logistic Regression for Weight Change}
if(!require(nnet)) {
  install.packages("nnet")
  library(nnet)
}

# Ensure 'weight_change' has a clear reference category
# Find the most common weight change category to use as reference
most_common <- names(sort(table(analysis_data$weight_change), decreasing = TRUE))[1]
analysis_data$weight_change <- relevel(factor(analysis_data$weight_change), ref = most_common)

# Build multinomial model
multi_model <- multinom(weight_change ~ Time.spent.on.sleep + Time.spent.on.fitness + 
                        Time.spent.on.TV + Number.of.meals.per.day,
                     data = analysis_data)

summary(multi_model)

# Calculate p-values (not provided directly by multinom)
z_scores <- summary(multi_model)$coefficients / summary(multi_model)$standard.errors
p_values <- (1 - pnorm(abs(z_scores), 0, 1)) * 2
print(p_values)
```

## [4] VISUALIZATIONS FOR REPORT

```{r Visualizations for Report}
# Create effect plots for logistic regression
# Use the correct column names
predicted_probs <- data.frame(
  Time.spent.on.sleep = seq(min(analysis_data$Time.spent.on.sleep, na.rm = TRUE),
                           max(analysis_data$Time.spent.on.sleep, na.rm = TRUE),
                           length.out = 100)
)
# Add mean values for other predictors
predicted_probs$Time.spent.on.fitness <- mean(analysis_data$Time.spent.on.fitness, na.rm = TRUE)
predicted_probs$Time.spent.on.TV <- mean(analysis_data$Time.spent.on.TV, na.rm = TRUE)
predicted_probs$Number.of.meals.per.day <- median(analysis_data$Number.of.meals.per.day, na.rm = TRUE)

predicted_probs$Health_Issue_Prob <- predict(logistic_model, 
                                           newdata = predicted_probs,
                                           type = "response")

# Plot effect of sleep on health issues
ggplot(predicted_probs, aes(x = Time.spent.on.sleep, y = Health_Issue_Prob)) +
  geom_line(color = "blue", size = 1.2) +
  theme_minimal() +
  labs(title = "Predicted Probability of Health Issues by Sleep Duration",
       x = "Hours of Sleep per Day",
       y = "Probability of Reporting Health Issues")

# Create stacked bar chart for meal counts by health issue
ggplot(analysis_data, aes(x = factor(Number.of.meals.per.day), fill = has_health_issue)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(title = "Proportion of Health Issues by Number of Meals",
       x = "Number of Meals per Day",
       y = "Proportion",
       fill = "Health Issue Reported")
```

## [5] TABLES FOR REPORT

```{r Tables for Report}
# Create summary table of means for time variables by health status
time_summary <- analysis_data %>%
  group_by(has_health_issue) %>%
  summarise(across(all_of(time_vars), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = all_of(time_vars),
               names_to = "Activity",
               values_to = "Average_Hours")

# Format for nicer display
time_summary$Activity <- gsub("\\.", " ", time_summary$Activity)
time_summary$Activity <- gsub("Time spent on ", "", time_summary$Activity)
time_summary$Average_Hours <- round(time_summary$Average_Hours, 2)

# Create a summary table for the report
kable(time_summary, caption = "Average Time Spent on Activities by Health Status")

# ANOVA table for report
sleep_anova <- anova(anova_results[["Time.spent.on.sleep"]])
fitness_anova <- anova(anova_results[["Time.spent.on.fitness"]])

kable(sleep_anova, caption = "ANOVA: Sleep Duration by Health Status")
kable(fitness_anova, caption = "ANOVA: Exercise Time by Health Status")

# Logistic regression results table
log_summary <- summary(logistic_model)$coefficients
log_summary <- cbind(log_summary, odds_ratio_table$OR)
colnames(log_summary)[4] <- "Odds_Ratio"
kable(log_summary, digits = 3, caption = "Logistic Regression: Predicting Health Issues")

# Save key tables and plots for report
write.csv(time_summary, "time_by_health_summary.csv")
write.csv(log_summary, "logistic_regression_results.csv")
```